---
title: "DASC 600 Assignment 04: Common Discrete Probability Distributions"
author: "Robert Johnson"
date: "`r format(Sys.time(), '%d %B %Y')`"
number-sections: true
number-depth: 3
format:
  html:
    toc: true
    toc-location: right
    number-sections: true
    number-depth: 3
    html-math-method: katex
    embed-resources: true
    self-contained: true
# bibliography: dasc-6000.bib 
# csl: ieee-with-url.csl
# linkcolor: red
# urlcolor: blue
# link-citations: yes
# header-includes:
#   - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Assignment goal

The overarching goal for this assignment is to assess your understanding of discrete probability distributions - Bernoulli, Binomial, Geometric, Negative Binomial, Poisson, Hypergeometric, Benford, Zipf, and mixtures.

# Instructions

Please show all your work. Simply providing the final answer is treated as no response. If you do not use R or Python notebooks, it is okay. Please preserve the document structure if you choose to use Microsoft Word or something else. Please number your responses to match with the corresponding questions.

**Please submit your response as a self-contained HTML or PDF document**. 

# Binomial Random Variable


1. A binomial random variable $X$ has population mean 3 and population variance $2.$ Compute $\mathbf{P}(X \leq 2)$. Also, verify your answer by computing $\mathbf{P}(X \leq 2)$ using R.

    
Mean tells us $np = 3$ and the variance gives us $np(1-p) = 2$
- Step 1: Substitute np and solve for p in variance
$$3(1-p) = 2 \rightarrow 3 - 3p = 2 \rightarrow 3p = 1 \rightarrow p = \frac{1}{3}$$
-Step 2: Substitute for p and solve for n in mean
$$n \cdot \frac{1}{3} = 3 \rightarrow n = 9$$
-Step 3: Compute probability for $X \leq 2$
```{r}
n <- 9 #n computed above
p <- 1/3 #p computed above
px <- pbinom(2, size=n, prob=p) #compute probability for 0-2
print(paste("P(X<=2) =", px))
```


# PowerBall Lottery

One version of the PowerBall lottery requires matching the numbers on five white balls, in any order, selected at random and without replacement from a drum containing 69 white balls numbered $1,2, \ldots, 69$, and matching an additional red "powerball" number selected at random from a second drum containing 26 red balls numbered $1,2, \ldots, 26.$

(a) What is the probability that one person who selects six numbers wins the PowerBall lottery?

The probability is equal to $\frac{1}{\text{Red Combinations} \cdot \text{White combinations}}$ or $\frac{1}{\binom{69}{5} \cdot \binom{26}{1}}$
```{r}
white_combos <- choose(69,5)
red_combos <- choose(26,1) #This clearly equals 26 but I include it for posterity
p <- 1 / (white_combos * red_combos)
print(paste("Probability of one person winning =", p))
```

(b) If one thousand people purchase PowerBall lottery tickets simultaneously and each person makes a random selection for the six numbers, independent of any of the other gamblers, find the probability mass function of the number of winners.

The pmf for this problem is given as $P(X=k)=\binom{n}{k} p^k (1-p)^{n-k}$

# Geometric Random Variable
    
Let $X$ denote the number of tails prior to the first occurrence of a head in repeated tosses of a fair coin. Find $P(X \bmod 5=2)$. 

The pmf for our problem is
$$P(X = k) = (\frac{1}{2})^k \cdot \frac{1}{2} = \frac{1}{2^{k+1}}, \text{for} \ k = 0,1,2,\ldots$$
We want the sum of the values that satisfy $X\ \text{mod} \ 5 = 2$
$$P(X \ \text{mod} \ 5 = 2) = \sum_{n=0}^{\infty} \frac{1}{2^{5n+3}}$$
Computing the seires takes a lot of steps but the final answer gotten is $\frac{4}{31}$

# Negative Binomial Random Variable

Let the random variable $X$ denote the number of independent rolls of a fair die required to obtain the second occurrence of a "five." Find $P(X \geq 4)$. 

```{r}
p <- 1/6 #p is for probability wherever you see it in here
successes <- 2
px2 <- choose(1,1) * (p^successes) * ((1-p)^(2-successes)) #P(X = 2)
px3 <- choose(2,1) * (p^successes) * ((1-p)^(3-successes)) #P(X = 3)
px4 <- px2 + px3 #P(X = 4)
pxg <- 1 - px4 #P(X >= 4)
print(pxg)
```


# Poisson Random Variable

An automobile manufacturer is implementing a "zero-defects" drive in order to improve the quality of their automobiles. They have found that the number of manufacturing defects on each car follows a Poisson distribution. If their goal is to have at least 98 percent of their cars defect-free, find the largest population mean $\mu$ that achieves their goal. 

Since we want zero defects our Poisson probability formula will have $k = 0$, and the goal is 98 percent of cars so we want $P(X=0) \geq 0.98$, thus it will take the form
$$P(X = 0) = \frac{\mu ^0 e^{-\mu}}{0!} = e^{-\mu} \geq 0.98$$

All that is left is to solve for $\mu$
```{r}
mu <- -log(0.98)
print(paste("Mu =", mu))
```

# Another Poisson Random Variable

The number of parking tickets issued daily is a Poisson random variable with population mean $15.$ Write the mathematical expression and use $R$ to find the probability that:

(a) exactly 17 tickets will be issued in one day,

$$P(X=17) = \frac{\lambda^{17}e^{-\lambda}}{17!} = \frac{15^{17}e^{-15}}{17!}$$
```{r}
lambda <- 15
p <- dpois(17, lambda)
print(paste("P(X = 17) =", p))
```

(b) 17 or more tickets will be issued in one day,

$$P(X \geq 17) = 1 - P(X \leq 16) = 1 - \sum_{k=0}^{16}\frac{\lambda^ke^{-\lambda}}{k!} = 1 - \sum_{k=0}^{16}\frac{15^k e^{-15}}{k!}$$
```{r}
p <- ppois(16, lambda, lower.tail=FALSE)
print(paste("P(X >= 17)", p))
```

(c) exactly 17 tickets will be issued in one day given that the population mean number of tickets was exceeded on that particular day, and

$P(X = 17)$ was defined in part (a)
$$P(X > 15) = 1 - P(X \leq 15) = 1 - \sum_{k=0}^{15}\frac{15^k e^{-15}}{k!}$$
$$P(X = 17 | X > 15) = \frac{P(X = 17 \cap X>15)}{P(X > 15)} =\frac{P(X = 17)}{P(X > 15)}$$
```{r}
px <- ppois(15, lambda, lower.tail=FALSE)
p <- dpois(17, lambda) / px
print(paste("P(X = 17 | x > 15) =", p))
```

(d) 100 or fewer tickets will be issued in any five-day period.

Let $\lambda_{Y}$ represent the average amount of tickets over 5 days then $\lambda_{Y} = 15 \cdot 5 = 75$
$$P(Y \leq 100) = \sum_{k=0}^{100} \frac{\lambda^k_Y e^{-\lambda_Y}}{k!} = \sum_{k=0}^{100} \frac{75^k e^{-75}}{k!}$$
```{r}
lambdaY <- 5 * lambda
p <- ppois(100, lambdaY)
print(paste("P(Y <= 100) =", p))
```

Assume that $Y$ be the random variable that represents the number of tickets issued in any five-day period.
    
Your answer goes here.
